---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit the Rmd, then knit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dflasso)
```

```{r hero-setup, include = FALSE, message = FALSE}
problem <- capital_allocation_problem(max_weight = 0.5)
train <- dfl_data(subset(capital_allocation_demo, split == "train"),
                  features = starts_with("feat_"),
                  cost = realized_return, scenario = scenario,
                  element_id = asset_id)
fit <- dfl_fit(train$x, train$cost, train$scenario, problem = problem,
               element_id = train$element_id,
               control = dfl_control(seed = 1))
```

# dflasso

Feature selection for predict-then-optimise problems that chooses features for better decisions, not better predictions.

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/Mischa-Hermans/dflasso/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Mischa-Hermans/dflasso/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Docker image version](https://img.shields.io/docker/v/mischahermans/dflasso?sort=semver&label=docker)](https://hub.docker.com/r/mischahermans/dflasso)
<!-- badges: end -->

## What it does

In a predict-then-optimise problem the decision is made before the true costs are known. Costs are estimated from features, then optimised over. Take routing a delivery along the shortest path, where the route is chosen before the day's travel times are known. Flood depth sits near zero on most days, so a prediction-focused model drops it as unimportant; on the day a crossing floods, though, it decides the route, and a model without it routes a truck into the flooded crossing. The same idea fits any predict-then-optimise problem, shortest path, knapsack, allocation or assignment, or a custom solver for one of your own.

The feature that predicts cost best and the feature that changes the chosen decision might be different ones. An ordinary lasso keeps the strong predictors and tends to drop the ones that change the decision but predict cost weakly. dflasso tries to include those variables in the model as well, by lowering the penalty on the features that move the decision most, so the refit is more likely to select them. It then solves the problem on held-out data and compares its regret with a prediction-focused model's. Regret is how much worse a decision turns out than the best possible in hindsight; lower is better.

The default plot shows this split directly: every feature placed by how well it predicts cost (horizontal axis) and how much it changes the decision (vertical axis). Features top-left are weak predictors that still change the decision; an ordinary lasso drops these, and dflasso marks the ones it keeps as rescued. Strong predictors that leave the decision unchanged sit bottom-right.

```{r hero, echo = FALSE, message = FALSE, fig.alt = "Features plotted by how well they predict cost (horizontal) and how much they change the decision (vertical); two weak predictors sit at the upper left and are kept, two strong predictors sit at the lower right, and two further features sit at the lower left and are left out."}
plot(fit)
```

## How it works

1. Fit a lasso cost model on the training instances, where one instance is a single optimisation problem.
2. Score each feature by how strongly its values track the held-out regret.
3. Ease the penalty on the features that move the decision most.
4. Refit the cost model under the eased penalties.
5. Solve each held-out instance with the new predicted costs.
6. Score regret against a prediction-focused baseline on the same instances.

Step 2 is the heavy one. dflasso resamples the instances, 15 times by default, and on each resample refits the lasso, re-solves the held-out instances, and correlates every feature with the resulting regret. That correlation runs in compiled C++, and the resampling loop runs in parallel when asked, through `dfl_control(parallel = TRUE, workers = 8)`. The built-in shortest-path solver and the knapsack dynamic program are compiled C++ as well; capital allocation uses lpSolve.

## Installation

Install from GitHub with pak:

```r
pak::pak("Mischa-Hermans/dflasso")
```

Or with remotes:

```r
remotes::install_github("Mischa-Hermans/dflasso")
```

Either call installs dflasso and the R packages it needs (glmnet for the lasso, lpSolve for the built-in solvers, ggplot2 for the plots), so there is no need to install those separately; DESCRIPTION lists the full set of libraries. dflasso needs R 4.1 or newer, and installing from source compiles C++, so build tools are needed: Rtools on Windows, the Xcode command-line tools on macOS, build-essential on Linux. Check for them with `pkgbuild::check_build_tools()`.

A Docker image is also available, as an optional alternative to installing in R, with dflasso and its dependencies already installed, including the compiled C++, so no build tools are needed:

```bash
docker pull mischahermans/dflasso:0.1.0
docker run --rm -it mischahermans/dflasso:0.1.0 R
```

## Quick start

`capital_allocation_demo`, one of two datasets bundled with the package, splits a budget across assets. The other, `aid_routing`, is the routing graph the vignette works through. The allocation data is a single data frame, one row per asset in each scenario, so the example runs without reshaping; its first instance, six assets, looks like this:

```{r data-peek, message = FALSE}
head(capital_allocation_demo[, c("scenario", "asset_id", "feat_01", "feat_02", "realized_return", "split")])
```

```{r quick-start, message = FALSE, fig.alt = "Features plotted by prediction importance (horizontal) and decision importance (vertical); two weak predictors sit high on decision importance and are kept."}
problem <- capital_allocation_problem(max_weight = 0.5)

train <- dfl_data(subset(capital_allocation_demo, split == "train"),
                  features = starts_with("feat_"),
                  cost = realized_return, scenario = scenario,
                  element_id = asset_id)

fit <- dfl_fit(train$x, train$cost, train$scenario, problem = problem,
               element_id = train$element_id,
               control = dfl_control(seed = 1))

fit
```

Each row is one element (here, one asset); the `scenario` column groups rows into an instance, which is one optimisation problem. See `?dflasso-glossary`. To use other data, supply one row per element, a scenario column grouping rows into instances, and a realised-cost column; see `?dfl_data` for the shape. For the problem, pick a constructor (`capital_allocation_problem()`, `shortest_path_problem()`, `knapsack_problem()`) or wrap a custom solver with `optimization_problem()`.

## Outputs

The calls below continue from the `fit` built in Quick start.

`summary(fit)` reports the kept features, the eased penalties, and what is known about held-out regret.

```{r summary, message = FALSE}
summary(fit)
```

`tidy(fit)` puts one row per feature into a tibble, with each feature's coefficient, decision-relevance score and penalty, ready to sort, filter or join:

```{r tidy, message = FALSE}
tidy(fit)
```

The penalty plot shows how far the decision focus lowered the lasso penalty on each feature it eased; a longer line means a bigger drop. Dark blue marks the features the refit then kept, as easing a penalty does not guarantee the feature is kept.

```{r penalty, message = FALSE, fig.alt = "For each feature the decision focus eased, a line from its usual lasso penalty to the eased penalty; the features the refit kept are dark blue, the rest grey, and all move leftward to a lower penalty."}
plot(fit, type = "penalty")
```

The coefficient path traces each feature's effect as the lasso penalty relaxes, with the chosen model marked.

```{r path, message = FALSE, fig.alt = "Coefficient paths against the lasso penalty, strong at the left and relaxing to the right; the kept features are highlighted by colour and linetype, and a dashed line marks the chosen model."}
plot(fit, type = "path")
```

`decide()` turns features into weights on new data, and never reads costs. The training frame stands in for new data here, to keep the example short.

```{r decide, message = FALSE}
out <- decide(fit, train$x, train$scenario,
              element_id_new = train$element_id)
out
decisions(out)[["scenario_01"]]
```

`regret()` scores both models on the held-out test split, the instances the fit did not train on.

```{r regret-text, message = FALSE}
test <- dfl_data(subset(capital_allocation_demo, split == "test"),
                 features = starts_with("feat_"),
                 cost = realized_return, scenario = scenario,
                 element_id = asset_id)

score <- regret(fit, test$x, test$cost, test$scenario,
                element_id_test = test$element_id)
score
```

The demo is simulated, with two features that move the allocation while predicting return poorly. On other data the decision focus may help less, or not at all. The held-out `regret()` is how to tell.

The boxes show how the per-instance regret is spread around each average.

```{r regret-plot, message = FALSE, fig.alt = "Per-instance regret of the decision-focused and prediction-focused models on the same held-out scenarios, with each average marked; the decision-focused model's average sits further left, at lower regret."}
plot(score)
```

## Beyond the built-ins

The package is not limited to its built-in problems. Any optimiser that turns predicted costs into a decision can be wrapped with `optimization_problem()` and used in place of a built-in problem. A model already written in lpSolve, igraph, ROI or a Python module can be wrapped rather than rewritten.

The solver is a single function, `function(costs, instance)`: `costs` holds the predicted costs for one instance, one entry per element row in the original row order, and the function returns the decision as a numeric vector of the same length and order. The order must line up with the rows, and dflasso cannot check it. `sense` sets whether the objective is minimised or maximised.

For example, a custom binary linear program that picks the `k` cheapest elements, solved with lpSolve:

```{r custom-solver, message = FALSE}
cheapest_k <- optimization_problem(
  sense = "min",
  solve = function(costs, instance) {
    solution <- lpSolve::lp(
      direction = "min", objective.in = costs,
      const.mat = matrix(1, nrow = 1, ncol = length(costs)),
      const.dir = "=", const.rhs = instance$k, all.bin = TRUE
    )
    as.numeric(solution$solution)
  },
  name = "cheapest-k"
)

solve_decision(cheapest_k, costs = c(5, 1, 4, 2, 3), instance = list(k = 2))
```

`solve_decision()` is the quickest way to check a solver: the two cheapest of the five elements come back as a 0/1 vector in the original order. Anything the solver reads from `instance` reaches it only through `instances`, passed to `dfl_fit()`; `make_instances(scenario, k = 2L)` attaches `k` to every instance.

The fit is then the Quick start call with `problem = cheapest_k` and the instances added:

```r
fit <- dfl_fit(train$x, train$cost, train$scenario,
               problem = cheapest_k,
               instances = make_instances(train$scenario, k = 2L),
               element_id = train$element_id,
               control = dfl_control(seed = 1))
```

`summary()` reads as before. Because `decide()` and `regret()` re-solve, they need the same instances, passed as `instances_new =` and `instances_test =`; their decisions are now 0/1 picks rather than weights. See `?dflasso-solvers` for worked assignment, transportation and spanning-tree solvers, and the order rules that matter for grid-shaped problems.

## When regret is already known

A solver is only needed when dflasso must measure regret itself. When per-instance regret is already known, from a model that already runs, `dfl_score()` ranks the features by how strongly each tracks that regret, with no fit and no solver:

```{r supplied-regret, message = FALSE}
set.seed(6)
day_regret <- pmax(0, rnorm(45, mean = 4, sd = 2))
day <- rep(sprintf("day_%02d", seq_along(day_regret)), each = 3)
regret <- rep(day_regret, each = 3)

backtest <- data.frame(
  day = day,
  feat_demand  = regret + rnorm(length(day), sd = 2),
  feat_weather = rnorm(length(day)),
  feat_noise   = rnorm(length(day)),
  regret = regret
)

dfl_score(backtest, features = starts_with("feat_"), scenario = day, regret = regret)
```

`regret_from_objectives()` and `regret_from_decisions()` build that regret vector from objective values or decisions already recorded, and `dfl_fit(regret = ...)` fits the weighted lasso from supplied regret instead of solving for it.

## Learn more

- [Reference and articles](https://Mischa-Hermans.github.io/dflasso/): the pkgdown site.
- `vignette("humanitarian-routing", package = "dflasso")`: the full worked example, on humanitarian routing.
- `vignette("custom-solver", package = "dflasso")`: a custom solver, from the problem through fit, decide and regret.
- Help pages: `?dflasso-glossary` for the vocabulary, with `?dflasso-faq`, `?dflasso-troubleshooting` and `?dflasso-validation` alongside.

## Getting help

- [GitHub Issues](https://github.com/Mischa-Hermans/dflasso/issues): bug reports and feature requests.

---

Developed by **Mischa Hermans**, Research Master's student in Econometrics and Operations Research at Maastricht University.
