---
output: github_document
---

<!-- README.md is generated from README.Rmd. Edit the Rmd, then knit. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(dflasso)
```

```{r hero-setup, include = FALSE, message = FALSE}
problem <- capital_allocation_problem(max_weight = 0.5)
train <- dfl_data(subset(capital_allocation_demo, split == "train"),
                  features = starts_with("feat_"),
                  cost = realized_return, scenario = scenario,
                  element_id = asset_id)
fit <- dfl_fit(train$x, train$cost, train$scenario, problem = problem,
               element_id = train$element_id,
               control = dfl_control(seed = 1))
```

# dflasso

Feature selection for predict-then-optimise problems that keeps the features that change the decision.

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/Mischa-Hermans/dflasso/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Mischa-Hermans/dflasso/actions/workflows/R-CMD-check.yaml)
[![License: MIT](https://img.shields.io/badge/license-MIT-blue.svg)](https://opensource.org/licenses/MIT)
[![Docker image version](https://img.shields.io/docker/v/mischahermans/dflasso?sort=semver&label=docker)](https://hub.docker.com/r/mischahermans/dflasso)
<!-- badges: end -->

## What it does

In a predict-then-optimise problem the decision is made before the true costs are known. Costs are estimated from features, then optimised over. Take splitting a budget across assets, where the weights are set before the returns are known: a feature can shift where the money goes while barely improving the return forecast. The same idea fits any predict-then-optimise problem, such as shortest path, knapsack, allocation, or assignment, with a custom solver for anything else.

The feature that predicts cost best and the feature that changes the chosen decision are often different ones, and dflasso keeps the second kind. An ordinary lasso ranks features by how well they predict cost. dflasso ranks them by how much they change the decision, eases the penalty on the ones that move the decision most, and refits. It then solves the problem on held-out data and compares its regret with a prediction-focused model's.

The default plot places every feature by how well it predicts cost (horizontal axis) and how much it changes the decision (vertical axis). Features top-left are weak predictors that still move the allocation; an ordinary lasso drops these, and the plot marks the ones dflasso keeps as rescued. Strong predictors that leave the decision unchanged sit bottom-right.

```{r hero, echo = FALSE, message = FALSE, fig.alt = "Features plotted by how well they predict cost (horizontal) and how much they change the decision (vertical); two weak predictors sit at the upper left and are kept, two strong predictors sit at the lower right, and two further features sit at the lower left and are left out."}
plot(fit)
```

## How it works

<details><summary>Show the steps</summary>

1. Fit a lasso cost model on the training instances.
2. Score each feature by how strongly its values track the out-of-sample regret.
3. Ease the penalty on the features that move the decision most.
4. Refit the cost model under the eased penalties.
5. Solve each held-out instance with the new predicted costs.
6. Score regret against a prediction-focused baseline on the same instances.

</details>

## Installation

dflasso is not on CRAN. Install the development version from GitHub:

```r
pak::pak("Mischa-Hermans/dflasso")
```

Or, with remotes:

```r
remotes::install_github("Mischa-Hermans/dflasso")
```

The Docker image includes dflasso and its dependencies; no R setup is needed:

```bash
docker pull mischahermans/dflasso:0.1.0
docker run --rm -it mischahermans/dflasso:0.1.0 R
```

Installing from source compiles C++, so build tools are needed: Rtools on Windows, the Xcode command-line tools on macOS, build-essential on Linux. Check it with `pkgbuild::check_build_tools()`.

## Quick start

`capital_allocation_demo` splits a budget across assets. It is a single data frame with one row per asset, so the example runs without reshaping.

```{r quick-start, message = FALSE, fig.alt = "Features plotted by prediction importance (horizontal) and decision importance (vertical); two weak predictors sit high on decision importance and are kept."}
problem <- capital_allocation_problem(max_weight = 0.5)

train <- dfl_data(subset(capital_allocation_demo, split == "train"),
                  features = starts_with("feat_"),
                  cost = realized_return, scenario = scenario,
                  element_id = asset_id)

fit <- dfl_fit(train$x, train$cost, train$scenario, problem = problem,
               element_id = train$element_id,
               control = dfl_control(seed = 1))

fit
```

Each row is one element (here, one asset); the `scenario` column groups rows into an instance, which is one optimisation problem. See `?dflasso-glossary`. To use other data, supply one row per element, a scenario column grouping rows into instances, and a realised-cost column; see `?dfl_data` for the shape. For the problem, pick a constructor (`capital_allocation_problem()`, `shortest_path_problem()`, `knapsack_problem()`) or wrap a solver with `optimization_problem()`.

## Outputs

The calls below continue from the `fit` built in Quick start.

`summary(fit)` reports the kept features, the eased penalties, and what is known about held-out regret.

```{r summary, message = FALSE}
summary(fit)
```

The penalty plot shows, for each feature whose filtering the decision focus eased, how far it was eased. A longer line means the penalty was eased more. Easing a feature's penalty does not guarantee the refit keeps it.

```{r penalty, message = FALSE, fig.alt = "For each feature whose filtering was eased, a line from its usual filtering strength to the eased, decision-focused strength; all move leftward, meaning easier to keep."}
plot(fit, type = "penalty")
```

The coefficient path traces each feature's effect as the model keeps more of them, with the chosen model marked.

```{r path, message = FALSE, fig.alt = "Coefficient paths for every feature against model size; the kept features are highlighted and a dashed line marks the chosen model."}
plot(fit, type = "path")
```

`decide()` turns features into weights on new data, and never reads costs. The training frame stands in for new data here, to keep the example short.

```{r decide, message = FALSE}
out <- decide(fit, train$x, train$scenario,
              element_id_new = train$element_id)
out
decisions(out)[["scenario_01"]]
```

`regret()` scores both models on the held-out test split, the instances the fit did not train on.

```{r regret-text, message = FALSE}
test <- dfl_data(subset(capital_allocation_demo, split == "test"),
                 features = starts_with("feat_"),
                 cost = realized_return, scenario = scenario,
                 element_id = asset_id)

score <- regret(fit, test$x, test$cost, test$scenario,
                element_id_test = test$element_id)
score
```

The demo is simulated, with two features that move the allocation while predicting return poorly. On other data the decision focus may help less, or not at all. The held-out `regret()` is how to tell.

The boxes show how the per-instance regret is spread around each average.

```{r regret-plot, message = FALSE, fig.alt = "Per-instance regret of the decision-focused and prediction-focused models on the same held-out scenarios, with each average marked; the decision-focused model's average sits further left, at lower regret."}
plot(score)
```

## Learn more

- [Reference and articles](https://Mischa-Hermans.github.io/dflasso/): the pkgdown site.
- `vignette("humanitarian-routing", package = "dflasso")`: the full worked example, on humanitarian routing.

## Getting help

- [GitHub Issues](https://github.com/Mischa-Hermans/dflasso/issues): bug reports and feature requests.
- A small [reprex](https://reprex.tidyverse.org/) makes a report quicker to act on.

## Citation

Hermans, M. (2026). dflasso: Decision-Focused Lasso for Predict-Then-Optimize Problems. R package version 0.1.0.

---

Developed by **Mischa Hermans**, Maastricht University.
