---
title: "A custom solver with dflasso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A custom solver with dflasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.width = 7, fig.height = 4.5)
library(dflasso)
```

A maintenance team can service three of its eight machines each week. It picks
the three with the lowest repair cost, but the cost is only known after the
work, so it is predicted from features beforehand. Some features predict a
machine's cost well; others barely predict it yet still decide which three are
cheapest that week. dflasso keeps the second kind.

No built-in problem fits "pick the cheapest three", so this vignette wraps one
as a custom solver and runs the whole workflow on it: fit the model, decide a
new week's machines, and check held-out regret against an ordinary
prediction-focused model.

## The decision, as a custom solver

`optimization_problem()` wraps any solver. This one takes the predicted costs
for a week's machines and returns a 0/1 vector marking the three to service.
The objective is `sense = "min"`, since lower cost is better, and the count `k`
arrives in `instance`.

```{r solver}
cheapest_k <- optimization_problem(
  sense = "min",
  solve = function(costs, instance) {
    chosen <- order(costs)[seq_len(instance$k)]
    decision <- numeric(length(costs))
    decision[chosen] <- 1
    decision
  },
  name = "cheapest-k"
)
```

The returned vector is the same length and order as `costs`. A larger problem
would call a solver such as lpSolve or igraph here instead; `?dflasso-solvers`
has worked examples.

## The data

Simulated repair records, one row per machine per week. Two week-level features
(load, temperature) set the week's cost level. Three per-machine features rank
the machines within a week: `feat_mileage` predicts cost well, while
`feat_fault` and `feat_wear` are weak predictors that still shift which three
are cheapest. `feat_noise` is unrelated.

```{r data}
simulate_repair <- function(n_weeks, seed) {
  set.seed(seed)
  n_machines <- 8L
  rows <- n_weeks * n_machines
  feat_week_load <- rep(rnorm(n_weeks), each = n_machines)
  feat_week_temp <- rep(rnorm(n_weeks), each = n_machines)
  feat_mileage <- rnorm(rows)
  feat_fault <- rnorm(rows)
  feat_wear <- rnorm(rows)
  feat_noise <- rnorm(rows)
  repair_cost <- 5 + 2 * feat_week_load + 2 * feat_week_temp +
    1.5 * feat_mileage + 0.2 * feat_fault + 0.2 * feat_wear +
    rnorm(rows, sd = 0.25)
  data.frame(
    week = rep(sprintf("week_%02d", seq_len(n_weeks)), each = n_machines),
    machine = rep(sprintf("machine_%d", seq_len(n_machines)), times = n_weeks),
    feat_week_load, feat_week_temp, feat_mileage,
    feat_fault, feat_wear, feat_noise, repair_cost,
    stringsAsFactors = FALSE
  )
}

train <- simulate_repair(80, seed = 1)
test  <- simulate_repair(40, seed = 2)
head(train[, c("week", "machine", "feat_mileage", "feat_fault", "repair_cost")])
```

`dfl_data()` slices the feature matrix, the cost, the week (the scenario that
groups a week's machines into one instance), and the machine id.

```{r prepare}
train_data <- dfl_data(train, features = starts_with("feat_"),
                       cost = repair_cost, scenario = week,
                       element_id = machine)
```

## Fit

`make_instances()` attaches `k = 3` to every week, so the solver knows how many
machines to service.

```{r fit}
fit <- dfl_fit(train_data$x, train_data$cost, train_data$scenario,
               problem = cheapest_k,
               instances = make_instances(train_data$scenario, k = 3L),
               element_id = train_data$element_id,
               control = dfl_control(seed = 1, n_splits = 10L))
summary(fit)
```

`feat_fault` and `feat_wear` are the rescues: weak predictors the decision step
kept because they move which machines are cheapest. `feat_mileage` and the
week-level features are kept the usual way, and `feat_noise` is dropped.

## Decide a new week's machines

`decide()` reads features only, never costs, so the choice computes before any
repair cost is known. The test weeks stand in for new weeks here.

```{r decide}
test_data <- dfl_data(test, features = starts_with("feat_"),
                      cost = repair_cost, scenario = week,
                      element_id = machine)
picks <- decide(fit, test_data$x, test_data$scenario,
                instances_new = make_instances(test_data$scenario, k = 3L),
                element_id_new = test_data$element_id)
decisions(picks)[["week_01"]]
```

The three 1s are the machines chosen for that week.

## Did the decision focus help?

`regret()` scores both models on the held-out test weeks: how much more the
chosen three cost than the cheapest three under the true costs, lower is better.

```{r regret}
score <- regret(fit, test_data$x, test_data$cost, test_data$scenario,
                instances_test = make_instances(test_data$scenario, k = 3L),
                element_id_test = test_data$element_id)
score
```

On this simulated data the decision focus roughly halves the held-out regret. It
keeps `feat_fault` and `feat_wear`, the two weak predictors the prediction-focused
model drops. The size is illustrative; on other data the focus may help less, or
not at all, and the held-out `regret()` is how to tell.

```{r spread, fig.alt = "Per-week regret for the decision-focused and prediction-focused models on held-out weeks; lower is better; each model's average is marked and the decision-focused one sits lower."}
plot(score)
```

## Where next

For a different problem, swap the solver inside `optimization_problem()`; the
fit, decide and regret calls stay the same. `?dflasso-solvers` has worked
solvers for assignment, transportation and spanning trees, and `?dfl_data`
covers the data shape.
