---
title: "Humanitarian routing with dflasso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Humanitarian routing with dflasso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE,
                      fig.width = 7, fig.height = 4.5)
library(dflasso)
```

An aid convoy is routed before the day's travel times are known, so they are
estimated from features and the fastest-looking path is taken. A main road's
time is easy to predict and rarely decides the route. Flood depth on a low
crossing is hard to predict, yet it flips the route the day the crossing
floods. A model tuned only for prediction keeps the main road and drops the
flood depth, then sends a truck onto the flooded crossing.

dflasso keeps the features that move the route, weak predictors included, and
refits. This vignette walks the whole loop on the `aid_routing` data: build the
rows, fit, decide tomorrow's route, and check on held-out days whether the
decision focus lowered regret against an ordinary prediction-focused model.

`aid_routing` is the package's demo dataset, and it is simulated: `flood_depth`
and `mud_depth` predict travel time poorly yet decide the route on wet days. So
read the method here, and take the magnitudes as illustrative. On other data the
regret check is how to tell.

## The problem, set up once

`shortest_path_problem()` is a built-in. It fixes the sense to "min" and solves
with Dijkstra, so no solver function is needed.

```{r problem}
routing <- shortest_path_problem()
```

## From graph tables to model rows

`aid_routing` is not a flat table. It is a list of graph objects: the arcs and
their per-arc features, the nodes, a day panel with each day's closures, and the
observed travel times. `prepare_instances()` joins them into the rows
`dfl_fit()`, `decide()` and `regret()` take, one row per open arc per day, and hands
back a coverage report. Each row is one element (here, one arc on one day), and the
scenario column groups a day's rows into an instance, the routing problem to
solve. See `?dflasso-glossary`. The `which` argument selects a slice of the
demo's day panel; on other data the rows are chosen directly.

```{r prepare}
train <- prepare_instances(aid_routing, which = "training")
dim(train$x)
head(train$coverage)
```

Read the report. Only the driven path is observed, so most (date, arc) pairs
never get a time and many days are set aside for scoring. Every observed time
trains the cost model; a day counts for scoring only when every arc the router
might use has an observed time, so most days are set aside by design. These few rows are clean, but across the full panel only about 45% of days
clear the bar for scoring.

## Fit it

One call with sensible defaults. With the seed named, re-runs reproduce the same
features and decisions. The fit here lowers `n_splits` from the default of 15 to
10 and runs sequentially so it knits fast; on a real graph leave the default and
add `workers = 8` for parallel splits. The
`element_id` argument is the per-arc label that ties model rows back to graph
arcs; `prepare_instances()` builds it, and the `_new` and `_test` forms are the
same label for `decide()` and `regret()` data. `instances` carries each day's graph for the solver: its open arcs (closed ones
already removed), the node count, and that day's origin and destination;
`prepare_instances()` builds it here.

```{r fit}
model <- dfl_fit(train$x, train$cost, train$scenario,
                 problem = routing,
                 instances = train$instances,
                 element_id = train$element_id,
                 control = dfl_control(seed = 2026, n_splits = 10L))
model
```

`summary()` is the one-screen report: how many features were kept, which weak
but decisive ones the method rescued, and the reproducibility fields. The
planted `flood_depth` and `mud_depth` are rescued, as intended.

```{r summary}
summary(model)
```

The proxy scores are correlational (see `?dflasso-validation`): read
`proxy_score` as the reliable signal and the rescue label as a rough summary.
With collinearity, and aggregation down to one number per scenario, a noise
column can clear the floor on another seed. The held-out `regret()` below
is the real test.

## Tomorrow's route

`decide()` needs features only, never costs, so tomorrow's route always
computes. Out comes the route, an ordered list of arcs.

```{r decide}
tomorrow <- prepare_instances(aid_routing, which = "tomorrow")
routes <- decide(model, tomorrow$x, tomorrow$scenario,
                 tomorrow$instances,
                 element_id_new = tomorrow$element_id)
element_sequence(routes)[["2026-08-12"]]
```

One planted day, `2026-08-14`, closes a cut-set that blocks every path from
origin to destination. dflasso marks it infeasible with a clear reason rather
than crashing, and every other day still routes.

```{r infeasible}
infeasible_reasons(routes)
```

## Does the decision focus lower regret?

This is the held-out check. Score realised regret on held-out days, which the
model never saw, against the prediction-focused baseline, the adaptive lasso
named by `baseline = "adaptive"`. Lower regret is better.

```{r regret}
test <- prepare_instances(aid_routing, which = "holdout")
result <- regret(model, test$x, test$cost, test$scenario,
                 test$instances, element_id_test = test$element_id,
                 baseline = "adaptive")
result
```

On this data the decision focus roughly halves the held-out regret. The exact
figure shifts a little between computers and software versions. It is not
guaranteed: on another graph the focus can lose, and dflasso reports that the
same way, with the verdict reading "RAISED regret".

## Does it hold across days, or only on average?

The mean can hide a few disastrous days. The paired spread plots per-day regret (one day is one instance here) for both
models, the printed average marked with a diamond.

```{r spread, fig.alt = "Per-instance regret for both models on held-out days; lower is better; each model's average marked by a diamond; the decision-focused model sits lower."}
plot(result)
```

## Where next

For the feature list alone, read `selected_features(model)` and skip `decide()`.
For a different problem (knapsack, allocation, assignment, an LP, or a custom
solver), copy a template from `?optimization_problem`; the fit, decide and
regret calls stay the same; what changes is `problem =` and the rows fed in. To
run on other data, supply what `prepare_instances()` produced for the demo: a
feature matrix `x`, the observed `cost`, a `scenario` grouping key, an
`element_id` label, and for a graph the per-instance `instances`; pair these
with a `problem =` from the constructors above. `make_instances()` builds those `instances` objects on other data, and
`?dfl_data` covers the
flat-table path.

```{r selected}
selected_features(model)
```
